Active Typechecking and Translation

### 2014-03-15 oopsla13/att-oopsla13.tex
removed stuff

%\documentclass[12pt]{article}
\documentclass[10pt]{sigplanconf}
% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage{listings}
\usepackage[usenames,dvipsnames]{color}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{graphicx}

\usepackage[noindentafter]{titlesec}
%\titlespacing{\section}{0pt}{*0.5}{*0.25}
%\titlespacing{\subsection}{3pt}{*1.5}{*0.25}
%\titlespacing{\subsubsection}{3pt}{*1.5}{*0.25}
%\titlespacing{\paragraph}{0pt}{*0.5}{*1}
\usepackage{textpos}
\input{macros}
\input{macros-lamalpha}
\newtheorem{theorem}{Theorem}%[document]
\newtheorem{lemma}{Lemma}
\newtheorem*{corollary}{Corollary}
\newtheorem*{definition}{Definition}

\begin{document}

%\conferenceinfo{POPL '13}{somewhere} 
%\copyrightyear{2013} 
%\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Active Type-Checking and Translation}

%\authorinfo{~}{~}{~}
\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
       {School of Computer Science\\
        Carnegie Mellon University}
        {\{comar, jonathan.aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract}
% Active Code Completion
% Active Typechecking and Translation
% Active Type-Directed Source Code Prediction
% Actively-Typed Parsing

Researchers and domain experts typically describe new language-based abstractions as extensions of some existing language, but most statically-typed languages are actually \emph{monolithic}: they do not give users the ability to specify the semantics of new core types and operators from within. This choice has led to a proliferation of mutually incompatible standalone languages, each built around a small collection of privileged constructs. 
An alternative to this \emph{language-oriented} approach is to work in an extensible programming language where the compile-time behaviors determining the functionality of new core constructs are specified and implemented modularly within user libraries. Designing such a \emph{library-oriented} extensibility mechanism for statically-typed languages that is both safe and expressive is non-trivial. % -- extensions must not weaken the global safety properties of the language, compromise the integrity of the compilation process, or interfere with one other, while remaining flexible enough to admit a range of useful constructs.
This paper introduces a mechanism called active type-checking and translation (AT\&T) that aims to address these  challenges. By relying upon type-level computation in a novel way, AT\&T admits user specification of a wide range of compile-time behaviors over a fixed grammar in a safe and modular manner. We discuss two points in the design space: (1) a simple calculus designed to distill the essential concepts and admit formal safety theorems, and (2) a fully-implemented language called Ace that we use to demonstrate the expressive power of AT\&T in several real-world domains, including scientific computing, security, functional programming and object-oriented programming.
\end{abstract}

@@

\section{Introduction}
Programming languages have historically been specified and implemented monolithically. To introduce new primitive constructs, researchers or domain experts have developed a new language or a dialect of an existing language, with the help of tools like domain-specific language frameworks and compiler generators \cite{fowler2010domain}. 
Unfortunately, taking a so-called {\it language-oriented approach} \cite{journals/stp/Ward94}, where different languages are used for different components of an application, can lead to problems at language boundaries: a library's external interface must only rely on  constructs that can be expressed in all possible calling languages. This means that specialized invariants cannot be checked statically, decreasing reliability and performance. It also often requires that developers generate verbose and unnatural ``glue'' code, defeating a primary purpose of specialized languages: hiding these low-level details from end-user developers.

Extensible programming languages promise to decrease the need for new standalone languages by providing more granular, language-based support for introducing new primitive constructs (that is, constructs that cannot be adequately expressed in terms of existing syntactic forms and primitive operations.) Developers would gain the freedom to choose those constructs that are most appropriate for their application domain and development discipline. Researchers would gain the ability to distribute new constructs for evaluation by a broader development community without requiring the approval of maintainers of mainstream languages, who are naturally risk-averse and uninterested in niche domains.

A significant challenge that faces language extensibility mechanisms is in maintaining the overall safety properties of the language and compilation process in the presence of arbitrary combinations of user extensions. The mechanism must ensure that basic metatheoretic and global safety guarantees of the language cannot be weakened, that extensions are safely composable, and that type checking and compilation remains decidable. The correctness of an extension itself should be modularly verifiable, so that its users can rely on it for verifying and compiling their own code. These are the issues that we seek to address in this work.

The approach we describe, {\it active type-checking and compilation} (AT\&T), makes use of type-level computation in a novel way. To review, in languages supporting type-level computation, the syntactic class of types is not simply declarative. Instead, it forms a programming language itself (the {\it type-level language}). Types themselves are one kind of value in this language, but there can be many others. To ensure the safety of type-level computations, {\it kinds} classify type-level terms, just as types classify expression-level terms. The simplest example of a language featuring type-level computation is Girard's System $\text{F}_{\omega}$ \cite{tapl}. In $\text{F}_{\omega}$, types have kind $\star$ and type-level functions have kinds like $\star\rightarrow\star$. A growing number of implemented languages now feature more sophisticated type-level languages (see Section 5).  
We emphasize that type-level computation occurs during compilation, rather than at run-time, because type-level terms that are used where types would normally be expected must be reduced to normal form before type-checking can proceed.

The organization and key contributions of this paper are:
\begin{itemize}
\item In Section 2, we develop a core calculus, $\lamA$, and give simple examples of language features that can be expressed with it. We formalize the compiler front-end and state several lemmas that lead to useful safety theorems for the compiler and language as a whole.  We show how AT\&T requires that the language provide a solution to a type-level variant of Wadler's expression problem \ref{Wadler}. 
\item In Section 3, we briefly introduce the Ace programming language, which is based fundamentally on an  elaboration of AT\&T that supports a richer set of syntactic forms and a variant of type inference. It uses object-oriented inheritance to solve the type-level expression problem. A number of practical extensions have been written using Ace, including a complete implementation of the OpenCL type system (based on C99) as a library.
\item In Section 4, we briefly describe another point in the design space, a language design we call Birdie. Birdie lifts an extension of the Gallina language, used by the Coq proof assistant, into the type level (leading to a language with dependent kinds). This additional complexity allows for full proofs of correctness for type-level specifications, and can allow proofs soundness of functional specifications against conventional inductive specifications. The expression problem is solved using a constrained formulation of open data types, rather than using object-oriented inheritance. 
\item In Section 5 we compare AT\&T to previous work on extensible languages and compilers, metaprogramming systems and formal specification languages and conclude in Section 6 with a discussion of future work.
\end{itemize}

\section{Type-Level Specifications in $\lamA$}
\subsection{Example: Natural Numbers in $\lamA$} % G\"{o}del's $T$ 
We begin with a simple calculus with no primitive notion of natural numbers, nor any more general notion of an inductive data type. We can, however, concretely specify both the static and dynamic semantics of natural numbers, including the natural recursor of G\"{o}del's $T$ \cite{tapl}, using type-level specifications. Let us begin in Figure 1 with the type $\tvar{nat}$ and its constructors, $\tvar{z}$ and $\tvar{s}$.
\begin{figure}
$\tfam{NAT}{\kunit}{
	\tlam{\tof{\tvar{self}}{\kType{\fvar{NAT}}}}{\titype{\dint}}}{(\\
	~~~~
	\topp{\tops{z}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tconst{\tden{\titerm{0}}{\tvar{self}}}}}}{
	\\~~~~
	\tops{s}{\tlam{\tof{\tvar{self}}{\kTypeBlur}}{\tOp{
		\tlam{\tof{\tvar{d1}}{\kDen}}{
		\\~~~~~~~~~\tifeq{\ttypeof{\tvar{d1}}}{\tvar{self}}{\tden{\tvalof{\tvar{d1}}}{\tvar{self}}}{\terr}}}}}}
	)}{\Theta}{{\textsf{let~}}{\text{nat}=\ttype{\tunit}{\fvar{nat}}}{\textsf{~in~}}{x}}$
\caption{Specifying natural numbers in $\lamA$. This term is referred to as $\psit{nat}$.}
\end{figure}
\begin{figure*}
\small
$$\begin{array}{rccl}
\textbf{expressions} 				&	e	&	::=	&	\evar{x} \pipe 
														\elam{\evar{x}}{\psi}{e} \pipe 
														\eapp{e_{1}}{e_{2}} \pipe
												\eop{\psitype}{op}{
															\splat{\psi}{1}{m}
														}{

														} \pipe
														\eop{\psitype}{op}{
															\splat{\psi}{1}{m}
														}{
  												    		\splat{e}{1}{n}
														} \\
									& 		&		& 	\\

\textbf{type-level specifications} 	& \psi 	& ::= 	& 	\tvar{t} \pipe 
														\tlam{\splatTwo{\tvar{t}}{{:}\kappa}{1}{n}}{\psi} \pipe 
														\tapp{\psi}{\splat{\psi}{1}{n}} \pipe
														\tifeq{\psi_{0}}{\psi_{1}}{\psi_{2}}{\psi_{3}} \\

\text{standard terms}	 			& 		& \pipe	& 	\tunit \pipe 
														\tpair{\psi_{1}}{\psi_{2}} \pipe 
														\tfst{\psi} \pipe 
														\tsnd{\psi} 
														\\

\text{type families}				&		& \pipe	& 	\tfam{fam}{\kappaidx}{\psirep}{\theta}{\Theta}{\psi}\pipe
														\tfamcase{\psi}{\phi}{\psi_0}{\psi_1}\\

\text{types} 						& 		& \pipe	& 	\ttypestd \pipe 
														\tidx{\psitype} \pipe
														\trepof{\psitype}
														\\

\text{operator definitions} 		& 		& \pipe & 	\tconst{\psi} \pipe 
														\tOp{\psi} \\

\text{denotations} 				& 		 & 	\pipe	&	\tden{\psiiterm}{\psitype} \pipe 
														\tvalof{\psiden} \pipe
														\ttypeof{\psiden} \pipe
														\terr \\

\text{internal language}		&		&	\pipe	&	\titerm{\mu} \pipe \titype{\delta}\\

\text{programs}					&		&	\pipe	&	\tprog{e}\\

\text{family specifications}	&	\phi		&	::= & \fvar{fam} \pipe \tfamSpecStd\\
\text{operator lists}		&	\theta	&	::= &	\topsempty \pipe 
												\topp{\theta}{\tops{op}{\psi}}\\

\textbf{kinds} 					& \kappa	&	::=	&	\karrow{\splat{\kappa}{1}{n}}{\kappa} \pipe 
												\kunit \pipe 
												\kpair{\kappa_{1}}{\kappa_{2}} \pipe 
												\kType{\eta} \pipe 
												\kTypeBlur \pipe
												\kOp{n} \pipe
												\kDenk{\eta} \pipe 
												\kDen \pipe 
												\kIType \pipe \kITerm \pipe
												\kProg 
												\\

\text{family signatures}				& \eta		& 	::=	& \kFamStd\\
\text{operator list signatures}			& \Theta	&	::=	&	\cdot \pipe \Theta; \tvar{op}: \kappa\\
											 							&		&		&	\\
\textbf{internal terms} 				& 	\mu	&	::=	&	\ivar{u} \pipe 
												\ilam{\ivar{u}}{\delta}{\mu} \pipe 
												\ifix{\ivar{f}}{\delta}{\mu} \pipe 
												\iapp{\mu_{1}}{\mu_{2}} \pipe 
												\iIfEq{\mu_{0}}{\mu_{1}}{\mu_{2}}{\mu_{3}}\\
							& 		& 	\pipe	&	\iintlit \pipe 
												\iop{\mu_{1}}{\mu_{2}} \pipe 
												\ipair{\mu_{1}}{\mu_{2}} \pipe 
												\ifst{\mu} \pipe
												\isnd{\mu} \pipe
												\iup{\psi} 
												\\
\textbf{internal types}			&	\delta	&	::=	&    \darrow{\delta_1}{\delta_2} \pipe
												\dint \pipe
												\dpair{\delta_1}{\delta_2} \pipe
												\dup{\psi}\\
%\text{cases}					&\tcasev 	&	::=	&	\caseZ{\splat{\kappa}{1}{n}}{\psi} \pipe 
%												\caseS{\tcasev}{\tcasev}\\
\end{array}$$
\caption{Syntax of $\lamA$. $\mathbb{Z}$ denotes integer literals and $\klabel{label}$ denotes any label.}
\end{figure*}
\begin{figure*}
\begin{mathpar}
\inferrule{
	\overbrace{\tentailst{\cdot}{\psiprog}{\kProg}}^{\text{\small kind checking}}{ }\\
	\overbrace{\tevalsX{\psiprog}{\tprog{e}}{\noprob}}^{\text{\small specification normalization}}{ }\\
	\overbrace{\translates{\cdot}{e}{\mu}{\ttypestd}}^{\text{\small verification \& translation}}{ }
}{
%	\compiles{\psiprog}{\mu}{\psiidx}{\kappa}{\psirep}
	\compiles{\psiprog}{\mu}{\ttypestd}
}
\end{mathpar}
\caption{Central compilation judgement of $\lamA$.}
\end{figure*}
%\noindent
%The \textsf{def} construct is the type-level analog of the standard \textsf{let} construct. Indeed, the entire term consists of type-level constructs. The definition of $\tvar{nat}$ says that it is a type indexed by a uniquely labeled unit (that is, it is a simple type) that will be represented in the internal language by integers. The constructor $\tvar{z}$ is declared as a constant, denoted by the value $0$ and assigned the type $\tvar{nat}$. The successor operator $\tvar{s}$ first checks its argument to ensure that it is a $\tvar{nat}$ (which requires first checking the index's kind to ensure that it has the correct label using the $\sf{idxcase}_{D}$ statement). If so, it returns a denotation that assigns the type $\tvar{nat}$ to an internal term that increments the value of the argument. Otherwise, the special denotation $\terr$ is returned. We will return to explain these constructs in more detail shortly.
%This term, which we will call $\psit{nat}$, is an open term because it has not been used to write a program yet. To use this specification to write a program, we substitute a \textsf{program} term containing an expression referring to these definitions for $\tvar{x}$. For example, the following program corresponds to the  number $1$: 
$$\psi_{1} := \subst{\tprog{
	\eop{\text{nat}}{s}{}{\eop{\text{nat}}{z}{}{}}
}}{\psi_{\text{nat}}}$$
%\noindent
%This program will be compiled to the internal term $1 + 0$ and assigned the type $\tvar{nat}$. This notion is formalized by the {\it central compilation judgement} of $\lamA$, given in Figure 3. In this case, we will be able to derive that
$$
\compiless{\psi_{1}}{1 + 0}{\tvar{nat}}
$$
%Compilation will not succeed for invalid terms, e.g.
%\begin{eqnarray*}
%\psi_{-} & := & \subst{\tprog{\econst{\tvar{s}}}}{\tvar{x}}{\psi_{\text{nat}}}\\
%\psi_{-}' & := & \subst{\tprog{\eop{\tvar{s}}{\elam{\evar{n}}{\tvar{nat}}{\eop{\tvar{s}}{\evar{n}}}}}}{\tvar{x}}{\psi_{\text{nat}}}
%\end{eqnarray*}
%Indeed, compilation with $\psi_{\text{nat}}$ as given above admits all well-typed expressions of the simply-typed lambda calculus with natural numbers and no others, compiling $\tvar{nat}$s to internal terms that evaluate to their integer representations.
%\subsection{$\lamA$}
\begin{figure*}
$\fbox{\inferrule{}{\tentailsXt{\psi}{\kappa}}}$~
~~~~$\fCtx ::= \cdot \pipe \fCtx, \fSpecStd$
~~~~$\tvarCtx ::= \cdot \pipe \tvarCtx, \tof{\tvar{t}}{\kappa}$
\begin{mathpar}
\small
\inferrule{
	\tof{\tvar{t}}{\kappa} \in \tvarCtx
}{
  \tentailsXt{\tvar{t}}{\kappa}
}~(\text{var}_\psi)

\inferrule{ }{
	\tentailsXt{\tunit}{\kunit}
}~(\text{unit}_\psi)

\inferrule{
	\tentailsXt{\psi_{1}}{\kappa_{1}}\\
	\tentailsXt{\psi_{2}}{\kappa_{2}}
}{
	\tentailsXt{\tpair{\psi_{1}}{\psi_{2}}}{\kpair{\kappa_{1}}{\kappa_{2}}}
}~(\times_\psi)

\inferrule{
	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
}{
	\tentailsXt{\tfst{\psi}}{\kappa_{1}}
}~(\text{fst}_\psi)

\inferrule{
	\tentailsXt{\psi}{\kpair{\kappa_{1}}{\kappa_{2}}}
}{
	\tentailsXt{\tsnd{\psi}}{\kappa_{2}}
}~(\text{snd}_\psi)

>>>>

%\documentclass[12pt]{article}
\documentclass[10pt]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.
\usepackage[usenames,dvipsnames]{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{mathpartir}
\usepackage{amsmath}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{graphicx}
\usepackage{url}
\newcommand{\TODO}[1]{{\color{red} #1}}
%\usepackage{float}
%\floatstyle{ruled}
%\newfloat{figure}{tp}{lop}
%\floatname{figure}{Listing}

%\usepackage[noindentafter]{titlesec}
%\titlespacing{\section}{0pt}{*0.5}{*0.25}
%\titlespacing{\subsection}{3pt}{*1.5}{*0.25}
%\titlespacing{\subsubsection}{3pt}{*1.5}{*0.25}
%\titlespacing{\paragraph}{0pt}{*0.5}{*1}
\usepackage{textpos}
\input{macros}
\input{macros-lamalpha}
\newcommand{\ATT}{AT\&T}
\newtheorem{theorem}{Theorem}%[document]
\newtheorem{lemma}{Lemma}
\newtheorem*{corollary}{Corollary}
\newtheorem*{definition}{Definition}

\usepackage{alltt}
\renewcommand{\ttdefault}{txtt}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\usepackage{listings}
\lstset{
  language=Python,
  showstringspaces=false,
  formfeed=\newpage,
  tabsize=4,
  commentstyle=\itshape,
  basicstyle=\ttfamily\footnotesize,
  morekeywords={lambda, self, assert, as},
  numbers=left,
  numberstyle=\footnotesize\color{gray}\textsf,
  xleftmargin=2em,
  stringstyle=\color{mauve}
}
\lstdefinestyle{Bash}{
    language={}, 
    moredelim=**[is][\color{blue}\bf\ttfamily]{`}{`},
}
\lstdefinestyle{OpenCL}{
	language=C++,
	morekeywords={kernel, __kernel, global, __global, size_t, get_global_id, sin}
}

\begin{document}
\title{Active Typechecking and Translation}
%
%\authorinfo{~}{~}{~}
\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
       {School of Computer Science\\
        Carnegie Mellon University}
        {\{comar, jonathan.aldrich\}@cs.cmu.edu}   
%
\maketitle
\begin{abstract}
Researchers and domain experts typically describe new language-based abstractions as extensions of an existing  language, but most statically-typed languages are \emph{monolithic}: they do not give users the ability to specify the static and dynamic semantics of new types and operators from within. This has lead to a proliferation of mutually incompatible standalone languages, each built around a small collection of privileged constructs. 
An alternative to this \emph{language-external} approach is to work in an extensible programming language where the compile-time behaviors determining the functionality of new core constructs are provided directly within user libraries. Designing such a \emph{language-internal} extensibility mechanism that is both safe  and expressive is non-trivial. % -- extensions must not weaken the global safety properties of the language, compromise the integrity of the compilation process, or interfere with one other, while remaining flexible enough to admit a range of useful constructs.
This paper introduces a mechanism called active type-checking and translation (AT\&T) that aims to address these  challenges. By building upon type-level computation in a novel way, AT\&T admits user specification of a wide range of compile-time behaviors over a fixed grammar in a safe and modular manner. We discuss two points in the design space: (1) a simple calculus designed to distill the essential concepts and admit formal safety theorems, and (2) a fully-implemented language called Ace that we use to demonstrate the expressive power of AT\&T across several domains, including scientific computing, security, functional programming and object-oriented programming.
\end{abstract}

@@

\section{Introduction}
Programming languages have historically been specified and implemented monolithically. To introduce new primitive constructs, researchers or domain experts have developed a new language or a dialect of an existing language, often with the help of tools like domain-specific language frameworks and compiler generators \cite{fowler2010domain}. 
Unfortunately, using different languages for different components of an application, called a {\it language-oriented approach} \cite{journals/stp/Ward94}, has problems when crossing language boundaries: the external interface of a library must only use constructs that can be expressed in all possible calling languages. \TODO{(talk about typical design using a common language like the JVM)} Thus, any specialized invariants cannot be checked statically, decreasing reliability. It also often requires that developers generate unnatural ``glue'' code, impacting performance and defeating a primary purpose of specialized languages: abstracting these low-level details away from developers.

Extensible programming languages promise to decrease the need for new standalone languages by providing more granular, language-internal support for introducing new primitive constructs (i.e. constructs that cannot be correctly and concisely expressed in terms of existing constructs). By using an extensible language, developers would gain the freedom to choose those constructs most suitable for their problem domain and development discipline. Researchers and domain experts would gain the ability to develop new constructs modularly and distribute them for evaluation by a broader development community without requiring either wholesale adoption of a new language or approval from the maintainers of widely-used languages, who are naturally risk-averse and are unlikely to cater to niche domains. The key difference between the language-external approach, characteristic of language frameworks,  and this language-internal approach, characteristic of extensible languages, is illustrated in Figure \ref{approaches}.
\begin{figure*}
\begin{center}
\includegraphics[scale=0.5]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages consisting of common constructs. (b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries. Thus, interoperability is not a problem. in monolithic systems, this approach is less expressive.\label{approaches}}
\end{figure*}

A number of criteria constrain potential language-internal extensibility mechanisms, however. One key issue is that a candidate mechanism must maintain the safety properties of the language and compilation process in the presence of arbitrary combinations of user extensions. The mechanism must ensure that basic metatheoretic and global safety guarantees of the language cannot be weakened, that extensions are safely composable, and that typechecking and compilation remain decidable. The correctness of extensions themselves should be modularly verifiable, so that developers can rely on them for verifying and compiling their own code. At the same time, the extensibility mechanism should be capable of expressing a variety of language constructs naturally (from the perspective of the developers using the constructs). For extension developers, using the extension mechanism should be comparable in difficulty to using DSL frameworks and other language-external tools. These criteria guide the work that we describe here.

We will begin by considering the typical compiler front-end, consisting of a typechecker and an initial translator to a high-level intermediate representation (e.g. bytecode for a virtual machine). The extensibility problem considered in this context can be cast as an instance of the \emph{expression problem} first examined by Reynolds \cite{reynoldsexpressionproblem} and so-named by Wadler \cite{expressionproblem}. Due to the particular structure of this domain, it is possible to sidestep the problem by delegating to functions associated with the internal representation of types, a form of \emph{inversion of control}.

Building an extensible compiler front-end by exposing these representations via some API is a tempting approach, but it remains a  \emph{language-external} one. That is, a compiler that allows for new language constructs (by any mechanism) creates, \emph{de facto}, a new language (an underappreciated fact in many language communities!)  Thus, programs compiled against different collections of extensions may be mutually incompatible with one another. Moreover, such extensions cannot be used by different compilers for the same language.

All is not lost, however. It can be observed that types are represented both within the compiler front-end and within the language itself. In most languages, the mechanisms for defining and referring to types are purely declarative. In languages featuring support for \emph{type-level computation}, however, types are a form of value that can be manipulated in various ways at compile-time via an internal \emph{type-level language} \cite{tlc}. This observation forms the foundation of the extensibility mechanism we describe here, called \emph{active typechecking and translation} (AT\&T), so named because it can be seen as a refinement of the general concept of \emph{active libraries}  proposed by veldHuizen \cite{activelibraries} (see Section \ref{relatedwork} for further discussion).

We will introduce AT\&T by way of two artifacts: a fully-implemented language, Ace, and a minimal type-theoretic formulation, $\lamA$. Ace demonstrates the practicality of AT\&T across a number of problem domains, beginning with low-level, performance-critical programming (targeting the OpenCL language and compilation infrastructure) and then demonstrate support for language-internal extensions that add a variety of higher-level  abstractions in Section \ref{extensions}. Ace, as an implemented prototype that makes some concessions for the sake of usability, does not readily admit formal safety theorems. For this reason, we introduce $\lamA$, a terse and somewhat more limited formalism based on a typed lambda calculus with type-level computation of higher kind. This calculus captures the core elements of the   mechanism and admits several important safety theorems.

%The Ace grammar is a nearly-complete subset of Python's  and we show that it is flexible enough to support all of the primitives of the C99 and OpenCL languages in a natural manner, when aided by a form of type inference that we describe. We further supplement these primitives with higher-level user-defined primitives, examples of which we discuss in Section \ref{abstractions}. Notably, all of these primitives are imported via the normal library loading mechanism; they are not defined within domain-specific languages, nor as extensions for a particular compiler implementation.
%

%The organization and key contributions of this paper are:
%\begin{itemize}
%\item In Section 2, we develop a core calculus, $\lamA$, and give simple examples of language features that can be expressed with it. We formalize the compiler front-end and state several lemmas that lead to useful safety theorems for the compiler and language as a whole.  We show how AT\&T requires that the language provide a solution to a type-level variant of Wadler's expression problem \ref{Wadler}. 
%\item In Section 3, we briefly introduce the Ace programming language, which is based fundamentally on an  elaboration of AT\&T that supports a richer set of syntactic forms and a variant of type inference. It uses object-oriented inheritance to solve the type-level expression problem. A number of practical extensions have been written using Ace, including a complete implementation of the OpenCL type system (based on C99) as a library.
%\item In Section 4, we briefly describe another point in the design space, a language design we call Birdie. Birdie lifts an extension of the Gallina language, used by the Coq proof assistant, into the type level (leading to a language with dependent kinds). This additional complexity allows for full proofs of correctness for type-level specifications, and can allow proofs soundness of functional specifications against conventional inductive specifications. The expression problem is solved using a constrained formulation of open data types, rather than using object-oriented inheritance. 
%\item In Section 5 we compare AT\&T to previous work on extensible languages and compilers, metaprogramming systems and formal specification languages and conclude in Section 6 with a discussion of future work.
%\end{itemize}

\section{Inversion of Control in a Compiler}\label{inversion}

\section{Structure and Usage}
%Ace is an {\em actively-typed, language-integrated compilation environment}. Let us begin by defining each of these terms. 
%A {\em compilation environment} is a system that allows developers to programmatically control aspects of compilation directly. The language used to do so is called the {\em metalanguage} while the grammar that programs themselves are written in is called the {\em source grammar}. The metalanguage of Ace is Python and the source grammar is a significant subset of Python's own grammar. Programs and metalanguage code can be interleaved within the same file, so Ace is {\em language-integrated}. Ace programs are statically-typed, and types themselves are objects in the metalanguage. They can be equipped by developers with methods that control how operators associated with them are typechecked and compiled. Because of the relationship between this mechanism and the idea of {\em active libraries} \cite{activelibraries}, discussed further in Section \ref{related}, we call them {\em active types}, and thus Ace is {\em actively-typed}.
\begin{figure}
\lstinputlisting{hello.py}
\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
\label{hello}
\end{figure}
\begin{figure}
\begin{lstlisting}[style=Bash]
$ `acec hello.py`
Hello, compile-time world!
Goodbye, compile-time world!
$ `cat hello.cl`
__kernel void main() {
    printf("Hello, run-time world.");
}\end{lstlisting}
\caption{Compiling \texttt{hello.py} using the \texttt{acec} compiler.}
\label{helloout}
\end{figure}

\subsection{Example 1: Hello, World!}
To demonstrate the somewhat unconventional structure of Ace programs and libraries, we begin in Listing \ref{hello} with a variant of the canonical ``Hello, World!'' example written using the \verb|OpenCL| module. We note that this module, which we use throughout the paper in examples, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed with Ace for convenience. Aspects of its implementation will be described in Section \ref{att}.

Perhaps the most immediately apparent departure from the standard ``Hello, World!'' example comes on lines 3 and 10 of Listing \ref{hello}, which contain \verb|print| statements that are executed at {\em compile-time}. Indeed, Ace programs are Python scripts at the top-level, rather than a list of declarations as in most conventional languages. In other words, Python is the {\em compile-time metalanguage} of Ace; Ace programs are embedded within Python. A consequence of this choice is that Ace can leverage Python's well-developed package system and  distribution infrastructure directly (Line 1).

The \verb|main| function defined on Lines 5-7 contains a single call to the \verb|printf| primitive, imported from the \verb|ace.OpenCL| module on Line 1. The function is introduced using the \verb|@OpenCL.fn| decorator, which indicates that it is a statically-typed Ace function containing run-time logic targeting the \verb|OpenCL| backend, also imported on Line 1. Without this decorator, the function would simply be a conventional Python function that could be called only at compile-time (doing so would fail here: \verb|printf| is not a Python primitive.)

On Line 8, this {\em generic function} defined on Lines 5-7, is compiled to produce a {\em concrete function} that we also identify as \verb|main| (overwriting the previous definition). The distinction between generic and concrete functions will be explained in our next example. For now, we note that only concrete functions are exported when running the Ace compiler, called \verb|acec|, shown operating at the shell in Listing \ref{helloout}. The \verb|acec| compiler operates as follows:
\begin{enumerate}
\item Executes the provided Python module (\verb|hello.py|)
\item Produces source code associated with any concrete functions (functions produced using the \verb|compile| method on Line 8 of Listing \ref{hello}) that are in the top-level environment. Each backend may produce one or more files (here, \verb|hello.cl| is produced).
\end{enumerate}

Using the \verb|acec| executable is simply a convenience; the generated code is also available as an attribute of the concrete function immediately after it has been defined, accessed as \verb|main.code|. We will show in Section \ref{pyopencl} that, for backend languages with Python bindings, such as OpenCL and C, generic functions can be executed directly, without the intermediate compilation step, if desired.

\subsection{Example 2: Higher-Order Map for OpenCL}
\begin{figure}
\lstinputlisting{listing3.py}
\caption{[\texttt{listing\ref{map}.py}] A generic data-parallel higher-order map function written using the OpenCL user module.}
\label{map}
\end{figure}

\begin{figure}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{mapadd5dbl}.py}] The generic \texttt{map} function compiled to map the \texttt{add5} function over two  types of input.}
\label{mapadd5dbl}
\end{figure}

\begin{figure}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{mapout}.cl}] The OpenCL code generated by running \texttt{acec listing2.py}.}
\label{mapout}
\end{figure}

The ``Hello, World!'' example demonstrates the basic structure of Ace programs, but it does not require working with types. Listing \ref{map} shows an imperative, data-parallel map primitive written using the OpenCL library introduced above. In OpenCL, users can define functions, called {\em kernels}, that execute across thousands of threads. Each kernel has access to a unique index, called its {\em global id}, which can be used by the programmer to ensure that each thread operates on different parts of the input data (Line 5). The \verb|map| kernel defined in Listing \ref{map} applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id. It writes the result of this call into the corresponding location in the provided output array, \verb|output|.

As above, \verb|map| is a {\em generic function} (specifically, an instance of the class \verb|ace.GenericFn|). This means that it's arguments have not been assigned types. Indeed, the functionality given by the \verb|map| definition is applicable to many combinations of types for \verb|input| and \verb|output| and functions, \verb|f|. In this sense, \verb|map| is actually a family of functions defined for all types assignments for \verb|input|, \verb|output| and \verb|f| for which the operations in the function's body are well-defined.

Running \verb|acec listing3.py| would produce no output, however. To create a {\em concrete function} (an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments of any externally callable functions. Listing \ref{mapadd5dbl} shows how to use the \verb|compile| method to specialize \verb|map| in two different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL has a notion of four different memory spaces). Line 9 produces a version specialized for arrays of \verb|double|s and Line 10 produces a version for arrays of \verb|int|s. The output of compilation is shown in Figure \ref{mapout}.

\subsection{Types as Metalanguage Objects}
The types of the arguments of the functions being compiled on Lines 4.9 and 4.10, if written directly in OpenCL as in Listing \ref{mapout}, are \verb|__global double*| and \verb|__global int*|, respectively. The corresponding types in Ace are \verb|global_ptr(double)| and \verb|global_ptr(int)|, abbreviated for convenience as \verb|A| and \verb|B| respectively on Line 8 (note that here, standard variable assignment in the metalanguage is serving the role that a \verb|typedef| declaration would fill in a C-like language.)

\textbf{Types in Ace are values in the metalanguage}, Python. More specifically, types are {\em object instances} of user-defined classes that inherit from the Ace-provided \verb|ace.Type| class. For example, \verb|global_ptr(double)| is an instance of \verb|ace.OpenCL.GlobalPtrType| instantiated with the target type, \verb|double|, as a constructor argument. The types \verb|double| and \verb|int| (imported from \verb|ace.OpenCL| on Line 4.2) are instances of \verb|ace.OpenCL.FloatType| and \verb|ace.OpenCL.IntegerType|, respectively. As we will describe further in Section \ref{att}, this notion of types as metalanguage objects is a key to Ace's flexibility.

\subsection{Type Propagation and Higher-Order Functions}
The type assigned to the third argument, \verb|f|, on Lines 4.9 and 4.10, is given as \verb|add5.ace_type|. The \verb|ace_type| attribute of a generic function is an instance of \verb|ace.GenericFnType|, the type of Ace generic functions, that corresponds to that particular function, \verb|add5| in this case. Generic functions are compiled automatically when they are called from another function. That is, when the compiler encounters the call to \verb|f| inside \verb|map| when compiling \verb|map_add5_double|, it compiles a version of \verb|add5| specialized to the \verb|double| type (seen on Line 5.3), and similarly with the \verb|int| type when compiling \verb|map_add5_int| (on Line 5.16, automatically given a unique name to avoid conflicts). This mechanism is called {\em type propagation}. In other words, we did not need to use \verb|add5.compile(double)| before compiling \verb|map_add5_dbl| (although we could if we would like). Only functions that are never called in the process of compiling other functions in a module need type information explicitly provided, whereas \verb|add5| is a function that is only called within \verb|map| in Listing \ref{mapadd5dbl}.
